---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r load_libs}
require("dplyr")
require("lubridate")
```

## Loading and preprocessing the data

Read the data from the [fork of the course git repository][1]. The repository was originally forked from [Reproducable Research Coursera repository][2]. As the data are read in, the "date" and "interval" variables are combined to create a "date.timstamp" variable.

```{r cache=TRUE}
activity <- tbl_df(read.csv("activity.csv")) %>%
  mutate(date.timestamp = ymd_hm(sprintf("%10s %04d", date, interval)))
```

## What is mean total number of steps taken per day?

Here is the frequency distribution of the total number of steps take each day of the study

```{r mean_and_median, cache=TRUE}
daily.steps.count <- activity %>% group_by(date) %>% summarize(step.cnt = sum(steps, na.rm = TRUE))
hist(daily.steps.count$step.cnt,
     xlab = "Total Daily Steps",
     ylab = "Frequency in Days",
     main = "Frequency Distribution of Daily Steps")
daily.steps.mean <- round(mean(daily.steps.count$step.cnt), 2)
daily.steps.median <- median(daily.steps.count$step.cnt)
```

The ***mean*** number of steps taken each day is **`r daily.steps.mean`** and the ***median*** is **`r daily.steps.median`** steps.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?

### Links
[1] https://github.com/davelpat/RepData_PeerAssessment1
[2] http://github.com/rdpeng/RepData_PeerAssessment1
